Word Sense Disambiguation: Pipelining with Supervised LSA

### 2014-05-13 paper.tex
completed paper

	\section{Introduction}
	\subsection{Problem Statement}
	Word sense disambiguation is the process of identifying the meaning of a word in its context when an individual word may take on many meanings.
	It remains one of the oldest open problems in computational linguistics and is widely considered to be AI-Complete, 
	or as hard as the problem of making computers as intelligent as people.$^1$ [CITATION] Being able to accurately disambiguate word senses would greatly
	assist Machine Translation, semantic analysis, and would have significant impact in computational linguistics. However, fine-grained senses
	in which a word has multiple closely related meanings remains one of the largest challenges in word sense disambiguation. Consider the following
	example of disambiguating the sense of the word {\emph {art}} from the Senseval-2 competition held in 2001:
	\begin{enumerate}
	\item "Paintings, drawings and sculpture from every period of {\emph {art}} during the last 350 years will be on display ranging from a Tudor portrait to contemporary British art."
	\item "Through casual meetings at cafe's, the artists drew together to form a movement in protest against the waste of war, against nationalism and against everything pompous, conventional or boring in the {\emph {art}} of the Western world."
	\end{enumerate}
	In the first example, {\emph {art}} refers to {\emph {the creation of beautiful or significant things}} whereas in the second example, it refers to {\emph {the products of human creativity; works of art collectively}}. Such subtle distinctions are part of what makes word sense disambiguation a challenging problem that remains open to this day.
	\section{Literature Survey}

        The past word sense disambiguation methods fall into three main categories: knowledge-based, unsupervised corpus-based, and supervised corpus-based methods.

@@

        One widely used knowledge-based algorithm is Lesk's algorithm.  The algorithm is as follows:

@@

	\subsection{Unsupervised Corpus-based}
	One of the challenges with word sense disambiguation is the scarcity of accurately sense-tagged text, which such datasets usually needing
	to be manually built by humans. Unsupervied approaches to word sense disambiguation attempt to overcome the bottleneck of acquiring sense-tagged
	text by operating on untagged corpora and clustering based on either context or word. Clustering by Committee (CBC) [CITATION] is an example of
	such an approach, formulated by Lin and Pantel in 2002, in which given a target word, clusters of similar words will be output. CBC is a
	three step algorithm starting first with the construction of a co-occurrence matrix containing measures of the similarity between each word,
	followed by clustering the most similar elements from this co-occrrence matrix using average link clustering to form the committees, with each
	target word then being assigned to its most similar committee in the final step.

	\subsection{Supervised Corpus-based}
	 Use annotated corpus to build model
 	 ex. to know: 1) be aware of piece of information
                              I want to know who is winning.
                              I know that the president lied.
                             2) know person
                             She doesn't know the composer.
                             Do you know my sister?
	 Use context of how word usually appears (bigram/trigram models)
 	 if (feature) then word sense
 	\subsection{LSA -- Past Efforts}
 	LSA was one of the first approaches to make use of the co-occurrence model for word meaning. The idea, as implemented in the famous 1998 paper by Landauer, is that given a set of corpuses, we build a term-document matrix where words are on the rows and documents are on the columns, and the frequencies are in the corresponding cells.
	Then we apply SVD to this matrix. 
	This approach is essentially unsupervised because it relies on clustering (word co-occurrence matrix) to derive similarities between documents and words.

@@

	(Katz et al) [CITATION] first applied LSA to WSD in an unsupervised approach by folding new documents with ambiguous word into the semantic space created by the SVD of the term-document matrix, and then used cosine-based clustering. We do a supervised implementation of this approach.

@@

	\subsubsection{Probabilistic LSA}
	Formulated as early as 1999 in a paper by Hofmann [CITATION], this approach is based on mixture decomposition from latent class model. It uses a probabilistic model known as an aspect model, which is a latent variable model for co-occurrence data. It introduces a conditional dependence assumption, namely that the document random variable and the word random variable are independent conditioned on the state of the associated latent variable.

@@

	Van de Cruys et al (2011) [CITATION] applies this approach to the SEMEVAL-2010 dataset. The rest of the approach is analogous to SVD-LSA, computing the centroid  of candidate vectors and comparing via Kullback-Leibler divergence to the target vector to determine word sense.

	\section{Data}
	Senseval-2 (2001) dataset
	73 words that need to be disambiguated
	         ex. fine, art, pull, turn, carry
	Up to 43 senses per word
	8611 labeled training instances
	4328 test instances

	brown corpus

@@

	 We detail the methods in the following sections.  
 	\subsection{Part-of-Speech}
 	We constructed two POS bigram models to try and perform coarse-filtering on the data by identifying the senses we could predict with high probability using a part-of-speech model before propogating the data instances further down the pipeline. We constructed one bigram model consisting of the POS bigrams containing the word coming before the target word and the target word, and another bigram model consisting of the POS bigrams containing the target word and word following the target word. The POS tags were generated by running NLTK's POS-tagger on the sentence containing the target word in each of our training data instances. On an input test instance, we would tag the sentence of interest, extract the two bigrams containing the target word, check if either bigram could be found in either of our bigram models, and output a predicted sense key if our predicted probability of the sense key given that bigram exceeded 80\%. To avoid skewed probabilities from having only one training data instance with that bigram, but also to ensure sufficient coverage for the already sparse data set, we also set a frequency count threshold of three such that we would only output our predicted sense key if we had seen more than three instances of this particular bigram in our training data and we predicted a probability of over 80\%.
 	\subsection{Semantic Knowledge base}
        \indent We attempted a couple of methods that were dictionary-based, but they yielded accuracy rates of $ < 30\%$, so we discarded them from the pipeline.

        We used Lesk's algorithm on the brown corpus, with an implementation that was based upon a previous implementation in python. [CITATIONS] We used a basic form of Lesk's algorithm that calculated dictionary paths. We also tried variations that involved including the dictionary definitions of homonyms to try to improve our accuracy, but it didn't change our results very much.

        We also used semantic similarity methods on the brown corpus, using an implementation by the same author as Lesk's algorithm. [CITATION] We tried all three methods: path similarity, Wu-Palmer, and Leacock-Chordorow's algorithm to see if they would yield high accuracy rates. However, each method was only accurate for certain words, but none of them could consistently identify the correct word sense.

@@

	\item Build term-document matrix 
		\begin{enumerate}
		\item terms are words in Brown Corpus and in training data
		\item documents are categories of Brown, and concatenated paragraphs of training data with same word sense for ambiguous word
		\end{enumerate}
	\item Apply SVD to get $U$, $D$, $V^T$.

@@

	We also tried a slightly different approach to capture more of the information: For each word, and for each sense of the word, we created a vector by applying cosine similarity to each eigentopic column vector for our training vector (created from paragraph). We averaged over all the training data for each word sense. Then for testing, we created this vector from the test paragraph and compared via cosine similarity again to determine which word sense was most likely. 
	This approach had issues and we got very bad results. 
	It might be better in the future to try a modified approach to this (i.e. do something other than average the vectors).
	\section{Results}
	From Senseval-2:

	Best overall (supervised approach): 64$\%$

	Best unsupervised approach: 40$\%$

@@

	Correctly disambiguates:
	"Marketing too, in its strictest sense, is outside our remit. " (the way in which a word or expression or situation can be interpreted)

@@

	a) With only the Brown Corpus categories as column vectors ($15$ column vectors $\rightarrow$ reduced to $9$) in the term-document matrix: $ ~ 42\%$
	b) With the Brown Corpus categories + the word-sense categories (in total, 861 column vectors $\rightarrow$ reduced to $140$ column vectors after SVD): $ ~ 38\%$
	-- highly dependent on the training data -- different results for different words, some words had lower percentages, some had higher percentages (as low as $7\%$, as high as $61\%$)
	-- easier to do ambiguous words with fewer senses

	\section{Conclusions}
	One issue with the LSA approach may be that the corpuses we use (namely, for a given word and a given sense, all the training paragraphs concatenated together) may actually cover a very broad array of topics, and thus a lot of noise might be added. While this should theoretically be removed by SVD, it might just be that the 'paragraph topics' were too disjoint and the correlation for different very fine-tuned usage is not strong enough. 
	Examples of failures:

	for the word "fine":

	our program: "elegant"  actual: "thin (in thickness)"
	our program: "superlative" actual: "elegant"

	and so on... it makes sense that both senses could apply to some of the testing data

	Labeled training data sparsity is a large issue

	Some statistical methods fall short of disambiguating subtle nuances between word senses

	The LSA approach is primarily topic categorization -- we need to use better documents that are more tailored to WSD, perhaps.

	It is hard to use a topic-categorization approach to do fine-tuned WSD.

	\section{Future Work}
	Could try different document columns for Supervised LSA.
	Could try applying supervised approach to PLSA or NNMF-LSA.

	Recall Socher et al. (2012) -- Recursive Neural Nets on sentiment trees
	Integrate with Socher's work on sentiment analysis
	Found that Socher doesn't disambiguate across certain words such as "mean", "like"
	Could improve accuracy of sentiment analysis

	Can be applied to translation if WSD is made better

	Maybe could apply SVM to the (word sense: cosine-vector) approach (i.e. train an SVM 
	for every ambiguous word in the training data with the cosine vectors as input)

@@

	1. Agirre, Eneko, and Philip Glenny. Edmonds. Word Sense Disambiguation: Algorithms and Applications. Dordrecht: Springer, 2006. Print.

	2. Phil Katz and Paul Goldsmith-Pinkham. 2006. Word Sense Disambiguation Using Latent Semantic Analysis Retrieved from http://www.sccs.swarthmore.edu/users/07 /pkatz1/cs65f06-final.pdf.

	3. T.K. Landauer, Foltz P.W., and D. Laham. 1998. Introduction to latent semantic analysis. $\textit{Discourse Processes}$, $25:259-284$.

	4. Thomas Hofmann. 1999. Probabilistic Latent Semantic Analysis. Retrieved from http://cs.brown.edu/~th/papers/Hofmann-UAI99.pdf.

	5. Tim Van de Cruys and Marianna Apidianaki. 2011. Latent Semantic Word Sense Induction and Disambiguation. Retrieved from http://aclweb.org/anthology//P/P11/P11-1148.pdf.

	6. Kirk Baker. 2013. Singular Value Decomposition Tutorial. Retrieved from http://www.ling.ohio-state. edu/~kbaker/pubs/Singular$\_$Value$\_$Decomposition$\_$Tutorial.pdf.

	8. Princeton University "About WordNet." WordNet. Princeton University. 2010. http://wordnet.princeton.edu.

    9. ACL-SIGLEX. 2001. Evaluation Exercises for the Semantic Analysis of Text. Retrieved from http://www.senseval.org/.

	10. R. Navigli. Word Sense Disambiguation: a Survey. ACM Computing Surveys, 41(2), ACM Press, 2009, pp. 1-69. [A complete State of the Art in Word Sense Disambiguation]

    11. Liling Tan. 2014. Pywsd: Python Implementations of Word Sense Disambiguation (WSD) Technologies [software]. Retrieved from https://github.com/alvations/pywsd.

>>>>

	\section{Introduction}
	\subsection{Problem Statement}
	Word sense disambiguation is the process of identifying the meaning of a word in its context when an individual word may take on many meanings.
	It remains one of the oldest open problems in computational linguistics and is widely considered to be AI-Complete, 
	or as hard as the problem of making computers as intelligent as people [2]. Being able to accurately disambiguate word senses would greatly
	assist Machine Translation, semantic analysis, and would have significant impact in computational linguistics. However, fine-grained senses
	in which a word has multiple closely related meanings remains one of the largest challenges in word sense disambiguation. Consider the following
	example of disambiguating the sense of the word {\emph {art}} from the Senseval-2 competition held in 2001 [1]:
	\begin{enumerate}
	\item "Paintings, drawings and sculpture from every period of {\emph {art}} during the last 350 years will be on display ranging from a Tudor portrait to contemporary British art."
	\item "Through casual meetings at cafe's, the artists drew together to form a movement in protest against the waste of war, against nationalism and against everything pompous, conventional or boring in the {\emph {art}} of the Western world."
	\end{enumerate}
	In the first example, {\emph {art}} refers to {\emph {the creation of beautiful or significant things}} whereas in the second example, it refers to {\emph {the products of human creativity; works of art collectively}}. Such subtle distinctions are part of what makes word sense disambiguation a challenging problem that remains open to this day.
	\section{Literature Survey}

        The past word sense disambiguation methods fall into three main categories: knowledge-based, unsupervised corpus-based, and supervised corpus-based methods [9].

@@

        One widely used knowledge-based algorithm is Lesk's algorithm [2].  The algorithm is as follows:

@@

	\subsection{Unsupervised Corpus-based}
	One of the challenges with word sense disambiguation is the scarcity of accurately sense-tagged text, which such datasets usually needing
	to be manually built by humans. Unsupervied approaches to word sense disambiguation attempt to overcome the bottleneck of acquiring sense-tagged
	text by operating on untagged corpora and clustering based on either context or word. Clustering by Committee (CBC) is an example of
	such an approach, formulated by Lin and Pantel in 2002 [2], in which given a target word, clusters of similar words will be output. CBC is a
	three step algorithm starting first with the construction of a co-occurrence matrix containing measures of the similarity between each word,
	followed by clustering the most similar elements from this co-occrrence matrix using average link clustering to form the committees, with each
	target word then being assigned to its most similar committee in the final step.

	\subsection{Supervised Corpus-based}
	 The supervised corpus-based methods use an annotated corpus to build a model, and makes predictions on previous word senses based on this model.  For example, for the definition to know, it could mean:

 	1) be aware of piece of information

    Usages: I want to know who is winning.  I know that the president lied.

    2) know a person

    She doesn't know the composer.

    Usages: Do you know my sister?

    The model uses the context of how a word usually appears and creates bigram/trigram models on surrounding words.  These models are applied to determine word sense in other instances.

 	\subsection{LSA -- Past Efforts}
 	LSA was one of the first approaches to make use of the co-occurrence model for word meaning. The idea, as implemented in the famous 1998 paper by Landauer [12], is that given a set of corpuses, we build a term-document matrix where words are on the rows and documents are on the columns, and the frequencies are in the corresponding cells.
	Then we apply SVD to this matrix. 
	This approach is essentially unsupervised because it relies on clustering (word co-occurrence matrix) to derive similarities between documents and words.

@@

	(Katz et al) [6] first applied LSA to WSD in an unsupervised approach by folding new documents with ambiguous word into the semantic space created by the SVD of the term-document matrix, and then used cosine-based clustering. We do a supervised implementation of this approach.

@@

	\subsubsection{Probabilistic LSA}
	Formulated as early as 1999 in a paper by Hofmann [10], this approach is based on mixture decomposition from latent class model. It uses a probabilistic model known as an aspect model, which is a latent variable model for co-occurrence data. It introduces a conditional dependence assumption, namely that the document random variable and the word random variable are independent conditioned on the state of the associated latent variable.

@@

	Van de Cruys et al (2011) [11] applies this approach to the SEMEVAL-2010 dataset. The rest of the approach is analogous to SVD-LSA, computing the centroid  of candidate vectors and comparing via Kullback-Leibler divergence to the target vector to determine word sense.

	\section{Data}
	We used the Senseval-2 dataset from 2001 [1], which consists of 73 words with multiple senses to be disambiguated. In total, there were 8611 training instances labeled with the WordNet sense key of the target word to be disambiguated, and 4328 test instances. All the data was provided in XML format. After parsing all training instances, we used only the sentence containing each target word for our POS bigram models and some of our dictionary-based methods, but the entire paragraph context for our LSA-based method. In addition to Senseval-2 dataset, we used the Brown corpus in our LSA-based method.

@@

	 We detail the methods in the following sections.  
 	\subsection{Part-of-Speech}
 	We constructed two POS bigram models to try and perform coarse-filtering on the data by identifying the senses we could predict with high probability using a part-of-speech model before propogating the data instances further down the pipeline. We constructed one bigram model consisting of the POS bigrams containing the word coming before the target word and the target word, and another bigram model consisting of the POS bigrams containing the target word and word following the target word. The POS tags were generated by running NLTK's POS-tagger [3] on the sentence containing the target word in each of our training data instances. On an input test instance, we would tag the sentence of interest, extract the two bigrams containing the target word, check if either bigram could be found in either of our bigram models, and output a predicted sense key if our predicted probability of the sense key given that bigram exceeded 80\%. To avoid skewed probabilities from having only one training data instance with that bigram, but also to ensure sufficient coverage for the already sparse data set, we also set a frequency count threshold of three such that we would only output our predicted sense key if we had seen more than three instances of this particular bigram in our training data and we predicted a probability of over 80\%.
 	\subsection{Semantic Knowledge base}
        \indent We attempted a couple of methods that were dictionary-based, but they yielded accuracy rates of $ < 30\%$, so we discarded them from the pipeline.

        We used Lesk's algorithm on the Brown corpus, with an implementation that was based upon a previous implementation in python. [5, 7] We used a basic form of Lesk's algorithm that calculated dictionary paths. We also tried variations that involved including the dictionary definitions of homonyms to try to improve our accuracy, but it didn't change our results very much.

        We also used semantic similarity methods on the brown corpus, using an implementation by the same author as Lesk's algorithm. We tried all three methods: path similarity, Wu-Palmer, and Leacock-Chordorow's algorithm to see if they would yield high accuracy rates. However, each method was only accurate for certain words, but none of them could consistently identify the correct word sense.

@@

	\item Build term-document matrix 
		\begin{enumerate}
		\item terms are words in Brown Corpus and in training data
		\item documents are categories of Brown, and concatenated paragraphs of training data with same word sense for ambiguous word
		\end{enumerate}
	\item Apply SVD to get $U$, $D$, $V^T$ [4].

@@

	We also tried a slightly different approach to capture more of the information: For each word, and for each sense of the word, we created a vector by applying cosine similarity to each eigentopic column vector for our training vector (created from paragraph). We averaged over all the training data for each word sense. Then for testing, we created this vector from the test paragraph and compared via cosine similarity again to determine which word sense was most likely. 
	This approach had issues and we got very bad results. 
	It might be better in the future to try a modified approach to this (i.e. do something other than average the vectors).
	\section{Results}
	For comparison, the highest accuracy obtained in the 2001 Senseval-2 competition was $64\%$ using a supervised approach. The best unsupervised approach had an accuracy of $40\%$.

@@

	Here are some examples of phrases that our POS Bigram model correctly disambiguates:
	"Marketing too, in its strictest sense, is outside our remit. " (the way in which a word or expression or situation can be interpreted)

@@

	We give two results: 
	with only the Brown Corpus categories as column vectors ($15$ column vectors $\rightarrow$ $9$ column vectors after SVD) in the term-document matrix, the accuracy of word sense disambiguation was $ ~ 42\%$ (average over all the words).

    With the Brown Corpus categories as well as the word-sense categories (in total, 861 column vectors $\rightarrow$ $140$ column vectors after SVD), the accuracy was on average $ ~ 31\%$, though individual accuracies for different words ranged from $7\%$ to $61\%$.

	\section{Conclusions}

	\subsection{Part-of-Speech Tagging}

	Part-of-speech performed typically very well on the subsets of the data that it deemed itself accurate enough to deal with. The main issue with POS was the subset of the data on which it was actually accurate enough to take care of. If the threshold was set at $80\%$, POS was applied to about only $6\%$of the data. Even when the threshold was set at $0\%$, POS only took care of $42.2\%$ of the data, since
	a large number of the bigram pairs (of part-of-speech relations) never showed up in the dataset. This indicates an issue with sparsity of the data.

	\subsection{Supervised LSA}
	One issue with the LSA approach may be that the corpuses we use (namely, for a given word and a given sense, all the training paragraphs concatenated together) may actually cover a very broad array of topics, and thus a lot of noise might be added. While this should theoretically be removed by SVD, it might just be that the 'paragraph topics' were too disjoint and the correlation for different very fine-tuned usage is not strong enough.

	Here are some examples of where our approach fails with Supervised-LSA:

	For the word "fine":

	Our program: "elegant"

	Actual: "thin (in thickness)"

	Our program: "superlative"

	Actual: "elegant"

	There are a large amount of errors in cases where it would be difficult for a human to disambiguate
	between the senses because the senses are very finely separated in meaning. Thus our approach falls short of disambiguating subtle nuances between word senses.

	Because the LSA approach is primarily topic categorization, we need to use better documents that are more tailored to WSD.

	Another issue is the sparsity of our labeled training data: because we potentially do not have enough 
	data for a given word sense, it becomes a lot more difficult to build a valid topic model. Thus it may be that we need more outside training data that has a single ambiguous word used in a specific word sense in multiple different ways. Another necessary improvement would be the use of this word sense in a constant topic area -- in other words, it would be nice if we had labeled data that used the word "fine" meaning "thin" only in sentences relating to cutting in a fine fashion (as opposed to being intermingled with sentences regarding a "fine figure", where "fine" still means "thin".)

	Again, since the accuracy was highly dependent on training data, it was easier to disambiguate
	ambiguous words with fewer senses because there were fewer possible ways to categorize the word senses.
	Also, the length of the document column texts were smaller as a result for the associated word senses.

	Due to the difficulty of gathering enough appropriate data, it is hard to use a topic-categorization approach to do fine-tuned WSD.

	\section{Future Work}

	There are many directions for future work with this project. 
	Regarding just the LSA-portion of the pipeline, several improvements could be made. We could try different document columns for Supervised LSA-- the method's success is in large part due to choosing document-columns appropriately. Perhaps it would make sense to create our own corpus and analyze a specific word to get really good accuracy with just disambiguating that word. We could also try improving the statistical rigor of the procedure itself by utilizing a supervised approach to PLSA or NNMF-LSA, as described earlier in the paper. We could also try an SVM approach with the supervised portion of the word -- build an SVM out of the cosine vectors, and train an SVM to recognize which word sense a cosine vector is most likely to be given a cosine vector for a test data input.

	The other approach we can take to extend this paper is applications. 
	Word-Sense Disambiguation would be relevant to the field of sentiment analysis. Socher et al (2013) [8]created a sentiment analysis pipeline with Recursive Neural Tensor Networks (RNTNs). However, there is no word-sense disambiguation step, which may affect the accuracy of the process, and words like "mean" and "like" are not disambiguated correctly. We could potentially improve the accuracy of sentiment analysis by including a word-sense disambiguation step (perhaps improved from ours) into the pipeline.

@@

    1. ACL-SIGLEX. 2001. Evaluation Exercises for the Semantic Analysis of Text. Retrieved from http://www.senseval.org/.

	2. Agirre, Eneko, and Philip Glenny. Edmonds. Word Sense Disambiguation: Algorithms and Applications. Dordrecht: Springer, 2006. Print.

	3. Bird, Steven, Edward Loper and Ewan Klein (2009), Natural Language Processing with Python. O’Reilly Media Inc.

	4. Kirk Baker. 2013. Singular Value Decomposition Tutorial. Retrieved from http://www.ling.ohio-state. edu/~kbaker/pubs/Singular$\_$Value$\_$Decomposition$\_$Tutorial.pdf.

    5. Liling Tan. 2014. Pywsd: Python Implementations of Word Sense Disambiguation (WSD) Technologies [software]. Retrieved from https://github.com/alvations/pywsd.

	6. Phil Katz and Paul Goldsmith-Pinkham. 2006. Word Sense Disambiguation Using Latent Semantic Analysis Retrieved from http://www.sccs.swarthmore.edu/users/07 /pkatz1/cs65f06-final.pdf.

	7. Princeton University "About WordNet." WordNet. Princeton University. 2010. http://wordnet.princeton.edu.

    8. Richard Socher, Alex Perelygin, et al. 2013. Recursive Deep Models for Semantic Compositionality Over a Sentiment Treebank. Retrieved from http://nlp.stanford.edu/$~$socherr/EMNLP2013$\_$RNTN.pdf.

	9. R. Navigli. Word Sense Disambiguation: a Survey. ACM Computing Surveys, 41(2), ACM Press, 2009, pp. 1-69. [A complete State of the Art in Word Sense Disambiguation]

	10. Thomas Hofmann. 1999. Probabilistic Latent Semantic Analysis. Retrieved from http://cs.brown.edu/~th/papers/Hofmann-UAI99.pdf.

	11. Tim Van de Cruys and Marianna Apidianaki. 2011. Latent Semantic Word Sense Induction and Disambiguation. Retrieved from http://aclweb.org/anthology//P/P11/P11-1148.pdf.

	12. T.K. Landauer, Foltz P.W., and D. Laham. 1998. Introduction to latent semantic analysis. $\textit{Discourse Processes}$, $25:259-284$.
